---
title: "Final Project"
author: ""
date: "2024-06-03"
output: pdf_document
---
#Data Collection
This data set is downloaded from Kaggle.
https://www.kaggle.com/datasets/aadarshvelu/aids-virus-infection-prediction

```{r}
library(tidyverse)
library(survival)
library(cluster)
library(ggplot2)
library(dplyr)
library(broom)
library(caret)
library(stats)
library(reshape2)

set.seed(333)

data <- read.csv("AIDS_Classification.csv")
```

Before we begin, we check if there happen to be missing values for this dataset
```{r}
cat("Number of missing values per column:\n")
sapply(data, function(x) sum(is.na(x)))
```
#Exploratory Data Analysis
There doesn't seem to be any missing data, now we can look at the frequencies of the classes to look at the distribution of the data
```{r}
#raw number of infect and noninfected
ggplot(data, aes(x=factor(infected), fill=factor(infected))) +
  geom_bar() +
  scale_fill_manual(values=c("green", "blue"), labels=c("Non-Infected", "Infected")) +
  labs(x="Infection Status", y="Number of People", fill="Status") +
  ggtitle("Histogram of Infected vs Non-Infected Individuals")

```
The data is not horribly imbalanced but still shows slight signs of imbalance. Checking the specific distributive percentages of the classes

```{r}
data_percentage <- data %>%
  group_by(infected) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100)

print(data_percentage)
```

##Visualizing key variables
```{r}
# visualize distributions of key variables
ggplot(data, aes(x=age)) + geom_histogram(bins=30, fill="blue") + ggtitle("Age Distribution")
ggplot(data, aes(x=wtkg)) + geom_histogram(bins=30, fill="green") + ggtitle("Weight Distribution")

# check for outliers
ggplot(data, aes(y=age)) + geom_boxplot(fill="coral") + ggtitle("Age Boxplot")
ggplot(data, aes(y=wtkg)) + geom_boxplot(fill="lightblue") + ggtitle("Weight Boxplot")

```

\vskip 0.5in
##Correlation Matrix
```{r}
data$infected <- as.numeric(data$infected)
numerical_data <- select_if(data, is.numeric)

correlation_matrix <- cor(numerical_data)

melted_correlation_matrix <- melt(correlation_matrix)

ggplot(melted_correlation_matrix, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(limit = c(-0.5, 1), mid = "white", high = "red", low = "blue", midpoint = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(title = "Correlation Matrix", x = "", y = "")
```
The vast number of variables make the correlation matrix hard to analyze.
\vskip 0.5in

##Feature Selection
Using statistical techniques to identify the most important features
```{r}
# using recursive feature elimination
control <- rfeControl(functions=rfFuncs, method="cv", number=10)
results <- rfe(data[, -ncol(data)], data[, ncol(data)], sizes=c(1:5), rfeControl=control)
print(results)

# Plotting feature importance
plot(results)

```

##model biulding
```{r}
# Logistic regression with glm
model <- glm(infected ~ age + time + cd40 + offtrt + preanti, data = data, family = binomial())
summary(model)

# extracting model coefficients
tidy_model <- tidy(model)
tidy_model$importance <- abs(tidy_model$estimate / tidy_model$std.error)

# plotting variable importance
ggplot(tidy_model, aes(x = reorder(term, importance), y = importance)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Variable Importance Plot", x = "Predictors", y = "Importance (|Coefficient/SE|)")

```
\vskip 0.5in

##clustering 
Lets begin by determining the #of clusters we want to use for k
```{r}
numeric_columns <- sapply(data, is.numeric)
data[numeric_columns] <- scale(data[numeric_columns])
new<- data[numeric_columns]
#Elbow method
wss <- sapply(1:10, function(k) sum(kmeans(data[numeric_columns], centers=k, nstart=10)$withinss))
plot(1:10, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares")
```
4 seems to be a reasonable choice

```{r}
# K-means clustering
kmeans_result <- kmeans(data[numeric_columns], centers=2, nstart=25)
data$cluster <- kmeans_result$cluster

# visualizing in pca
pca_results <- prcomp(data[numeric_columns])
pca_data <- data.frame(PC1 = pca_results$x[,1], PC2 = pca_results$x[,2], cluster = as.factor(kmeans_result$cluster))

# Assuming 'pca_data' contains the PCA results and cluster assignments
ggplot(pca_data, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha=0.7) +
  stat_ellipse(type = "t", linetype = 2, size = 1, level = 0.95) +  # Adds ellipses
  labs(title = "PCA of Dataset with K-Means Clusters", x = "Principal Component 1", y = "Principal Component 2") +
  scale_color_brewer(type = "qual", palette = "Set1") +
  theme_minimal()

#looking at loadings
loadings <- pca_results$rotation[, 1:2]  
print(loadings)

# Add cluster assignments to original data
data$cluster <- as.factor(kmeans_result$cluster)
summary_stats <- aggregate(. ~ cluster, data, mean)
print(summary_stats)

# Melting data for easier plotting
data_melted <- reshape2::melt(data, id.vars = "cluster")
ggplot(data_melted, aes(x = cluster, y = value, fill = cluster)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free_y") +
  theme_minimal() +
  labs(title = "Feature Distribution by Cluster", y = "Value", x = "Cluster")

```
The axes (PC1 and PC2) represent the principal components that account for the most variance in the data

##Hierarchical clustering

```{r}
hc_result <- hclust(dist(data[numeric_columns]), method="ward.D2")
plot(hc_result, main = "Dendrogram", xlab = "Index of Data Points", ylab = "Height",
     lty = 2, col = "blue", sub = "", cex = 0.6)
rect.hclust(hc_result, k = 4, border = "red")
```

\vskip 0.5in

##Do AIDS patients exhibit different patterns of treatment response based on the type of treatment received? How do CD4/CD8 counts change over time for patients under different treatment regimens?

Box plots 
```{r}
data <- read.csv("AIDS_Classification.csv")
data$trt <- as.factor(data$trt)

#CD4 counts at baseline
ggplot(data, aes(x = factor(trt), y = cd40, fill = factor(trt))) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = "CD4 Counts at Baseline by Treatment", x = "Treatment Group", y = "CD4 Count") +
  theme_minimal()

#D4 counts at 20 week
ggplot(data, aes(x = factor(trt), y = cd420, fill = factor(trt))) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = "CD4 Counts at 20 Weeks by Treatment", x = "Treatment Group", y = "CD4 Count") +
  theme_minimal()

# CD8 counts at baseline
ggplot(data, aes(x = factor(trt), y = cd80, fill = factor(trt))) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = "CD8 Counts at Baseline by Treatment", x = "Treatment Group", y = "CD8 Count") +
  theme_minimal()

#CD8 counts at 20 week
ggplot(data, aes(x = factor(trt), y = cd820, fill = factor(trt))) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = "CD8 Counts at 20 Weeks by Treatment", x = "Treatment Group", y = "CD8 Count") +
  theme_minimal()


```

Change plots
```{r}
data <- data %>%
  mutate(cd4_change = cd420 - cd40,
         cd8_change = cd820 - cd80)

# Change in CD4 and CD8 counts
ggplot(data, aes(x = factor(trt), y = cd4_change,fill = factor(trt))) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = "Change in CD4 Counts by Treatment from Baseline to 20 Weeks", x = "Treatment Group", y = "Change in CD4 Count") +
  theme_minimal()

ggplot(data, aes(x = factor(trt), y = cd8_change,fill = factor(trt))) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = "Change in CD8 Counts by Treatment from Baseline to 20 Weeks", x = "Treatment Group", y = "Change in CD8 Count") +
  theme_minimal()

```

Histogram
```{r}
#histogram for cd4-cd4 20 weeks
par(mfrow=c(1,2))
hist(data$cd40, main="Histogram of CD40", xlab="CD4 at baseline")
hist(data$cd420, main="Histogram of CD420", xlab="CD4 at follow-up")
par(mfrow=c(1,1))

#histogram for cd8-cd8 20 weeks
par(mfrow=c(1,2))
hist(data$cd80, main="Histogram of CD80", xlab="CD8 at baseline")
hist(data$cd820, main="Histogram of CD820", xlab="CD8 at follow-up")
par(mfrow=c(1,1))
```

ANOVA tables
```{r}
# ANOVA for CD4 counts from baseline to follow-up
anova_cd4 <- aov(cd40 ~ cd420 + trt, data = data)
summary(anova_cd4)

# ANOVA for CD8 counts from baseline to follow-up
anova_cd8 <- aov(cd80 ~ cd820 + trt, data = data)
summary(anova_cd8)

# ANOVA to test for differences in changes in CD4 counts across treatment groups
anova_result <- aov(cd4_change ~ factor(trt), data = data)
summary(anova_result)

```